### 2024-05-14 Manuscript/main.tex
references input

\documentclass[12pt,letterpaper]{article}

@@

In the last two decades, there has been a progressive shift in ecology and evolution from  taxonomic-oriented approaches, with species as the focal point of studies, to functional approaches that place species-specific characteristics (traits) at the centre of analyses (@MammolaFuncEcol2021, @Palacio2022).
Although numerous definitions exist (@Dawson), we here consider a trait to be any characteristic - e.g. , morphological, anatomical, ecological, physiological, behavioural, phenological—measured on individual organisms at any level, from genes to whole organisms.
A trait-based approach has two advantages for answering core questions in ecology and evolution.
First, it allows a deeper understanding of the mechanisms generating biodiversity patterns, by putting organisms’ traits at the centre of natural selection rather than organisms themselves.
Second, it allows comparisons across subdisciplines in biology, while facilitating the conceptualization of general principles broadly valid in space (e.g., unrelated species pools) and time (e.g., anatomical traits are comparable between palaeontology and ecology, which is not always the case with species) (@luza2023going).
Already used routinely in palaeontology (@Raup1961, @Gould1991, @Foote1995, @GuillermeBiolLet2020), this new trait-focused ecology and evolution is unlocking the possibility to answer a broad range of questions in disciplines as diverse as community ecology (@McGill), biogeography (@Violle), conservation biology (@Chichorro2022), micro- (@Chapin), macro-evolution (@Wiens), and applied fields (e.g. agronomy @Martin2015).
This is because trait-based approaches closely align with the general analytical framework proposed by @AnandCoenoses1994 to answer three consequential questions: "what?" (describing the pattern), "how?" (describing the process), and "why?" (understanding the mechanism) (Box 1).

\break

\hrule
\section*{Box}
@AnandCoenoses1994 conceptualised a general analytical framework for answering scientific questions in ecology and evolution, based on three consequential questions: "what?" (describing the pattern), "how?" (describing the process), and "why?" (understanding the mechanism).
This framework can be effectively applied to trait-based analyses:

\textit{What?} Pattern description corresponds to the steps needed to collect and summarise data to answer a question of interest (i.e., “how?” and “why?” as defined below).
Usually this consists in: i) collecting target traits at the focal level (e.g., genes, individual, population, species; @ViolleOikos2007); ii) arranging these traits into some kind of trait space (@GuillermeBiolLet2020, @MammolaFuncEcol2021); and iii) using one or more unidimensional statistics to summarise properties of the trait space.
For unidimensional spaces (i.e. distributions), these statistics can be as simple as the mean and the standard deviation (e.g.
 community weighted mean).
For multidimensional spaces, these are usually named disparity metrics or indices (@GuillermeBiolLet2020) or functional diversity metrics (@MammolaFuncEcol2021), but all attempt to capture some pattern of interest in the trait space.

\textit{How?} Process description can be seen as linking the pattern of interest (“What?”) to some dynamic element.
This can be a punctual change.
For example, change of the pattern under a certain condition, such as how traits X differ between two habitats (@Martinez).
But it can also be one or more continuous or ordinal changes, such as how the pattern X changes in space, time and/or along ecological gradients (@Belmaker, @Jaryna, @Lamanna, @bjorkman, @McLean).
The distinction here might seem trivial but it is significant: usually, the process designates the change of the pattern, not the change of the traits.
Although the traits are what is really changing, researchers will usually analyse some emergent property of the trait aggregation as described by the change of statistics between two or more conditions.

\textit{Why?} Mechanism description is then at the core of answering the biological question at hand.
In most cases, this is what researchers are actually trying to understand.
For example, one might be interested in understanding the effect of climate change on some trait (e.g. Does temperature affect leaf surface area?).
In fact, most researchers work on understanding the causal link between variables (i.e. “why” are “what” and “how” linked).
This can be very useful, for example, to provide predictions about the past or the future.
Ultimately, studying the mechanisms (“why””) is often the reason why researchers get funded (or not).

@@

\hrule

While the first part of this framework (“what?”) has been studied extensively in the last decade (@CITECITECITE), this was often undertaken outside of the “what, how, and why” framework.
Here we argue that the pattern description, although the basis of any future study, often follows some framework without a proper evaluation of its adequacy to answer the how and why.
The choice of the tool or metric used to describe the pattern is crucial for allowing us to understand the process and the mechanism.
Using an inappropriate metric for describing a pattern can lead to biased conclusions.
For example, let us imagine one is interested in understanding how two populations compete with each other for a resource (the mechanism - “why?”) on an island compared to another island with only one of the species (the process - “how?”).
In such a scenario, measuring the occupied area of each population on each island (a pattern - “what?”) will not be the most appropriate way to understand the potential competition between these populations.
In this simplistic example, the functional overlap between populations (“what and how”) might be more appropriate.
See for example Carvalho \& Cardoso (@ref) for a discussion of how Darwin’s finches share resources depending on the existence or not of competition.

Through a simulation exercise, we analysed different patterns (what) across different processes (how) approximating different mechanisms (why) of interaction between organisms: equalisation, filtering, facilitation and competition (Figure 1).
We also used an empirical dataset of Hawaiian bird traits and used pre-historical and historical extinction events on Hawaiian islands as patterns and processes capturing the effects of species extinction on species trait distributions.
We show that, with a fixed process and pattern, the choice of statistics used to describe the patterns (the disparity or functional diversity metrics) have a great impact on our ability to recover the process and understand the mechanism.
We suggest caution when summarising patterns in observed data and propose tools for researchers to help understand how their pattern (often stemming from multidimensional data) can vary intuitively or not depending on the process and mechanism of interest.

@@

First, we simulated a single Brownian Motion trait under a pure birth speciation model until reaching 200 elements (representing tips, individuals, species, OTUs, etc.).
This resulted in a neutral null model of trait evolution with no effect of competition, extinction, selection or other processes (sensu @BausmanPhilSci2018). We refer to this as the “non-stressed trait space”.

@@

We then applied five different stressors to the non-stressed trait space to remove either 20\%, 40\%, 60\% or 80\% of the data (resulting in trait spaces with, respectively, 160, 120, 80 and 40 elements).
In removing the data, we applied these five specific algorithms (Figure 1; Table 1; all algorithms, except “Evenness”, were previously described in @GuillermeEcoEvo2020):

@@

For the position, size and density algorithms, we estimated the parameters $\rho$ and $D$ recursively with the \texttt{dispRity} package to obtain the required amount of data to be removed (\texttt{dispRity::reduce.space}, Guillerme 2018, @GuillermeEcoEvo2020).
These steps resulted in 4000 simulated trait spaces (200 simulated spaces $\times$ 5 stressors $\times$ 4 amounts of data removed).
See Table 1 for a biological description of these stressors and Figure 1 for a visual description of them.

@@

We repeated each simulation pipeline (generate a traitspace and apply the stressor) for uncorrelated 4 dimensional traits (figure 2) and uncorrelated 2, 3, 5 and 8 dimensional traits (supplementary materials).
We limited our simulations to a relatively small number of traits due to the constraints of some of the metrics used (e.g., \texttt{TPD::TPDsMean} is only implemented for up to 4 dimensions) but also to avoid dealing with the curse of dimensionality.
This curse changes the properties of space occupancy in a non-linear way depending on the element's distribution and the number of dimensions.
For example, the volume of a trait space (or hypervolume when using more than 3 dimensions) typically tends to zero in a high number of dimensions.
However, the rate at which it approaches zero is not linear and depends on the distribution of each element on every dimension.
This makes it practically difficult to compare two randomly generated spaces with similar characteristics (e.g. for two spaces with 200 elements and 10 dimensions generated in the exact same way, one might have a hypervolume of nearly 0 and the other one of $10^5$).
Note also that in our simulations, all dimensions have the same properties (i.e.
same variance and distribution).
This is often not the case in empirical cases (e.g. see our empirical example) where the dimensions have a decreased variance due to ordination techniques.

\begin{table}
\center
\scriptsize
\begin{tabular}{p{0.3\linewidth}|p{0.3\linewidth}|p{0.4\linewidth}}
\textbf{Approximated mechanism (biological)} & \textbf{Used mechanism (algorithm in \texttt{dispRity::reduce.space})} & \textbf{Stressor description}\\
\hline
  Null mechanism & Random removal (“random”) & No overall change in species traits or community structure except for the reduction in the number of elements.\\
  Equalising fitness & Size change (“size”) & When species with trait combinations that are more extreme (i.e. species located away from the centre of the trait space) are disadvantaged due to some resource concentration gradient. \\
  Facilitation & Density change (“density”) & When species with closely shared trait combinations (i.e. located close together in the trait space) are more likely to survive the stressor. In other words, this simulates the idea that some species are more likely to be resilient if they are present in a community of species with shared traits rather than the opposite.\\
  Environmental filtering & Position change (“position”) & When an increase in species trait similarity happens through some strong abiotic constraints (@Cornwell2006). For example, some trait combinations become more and more unlikely due to some environmental constraints (i.e. some regions of the trait space become unsuitable).\\
  Competitive exclusion & Evenness change (“evenness”) & When functionally similar species compete more intensively with one another than with functionally dissimilar species (``competitive exclusion principle''; @Hardin 1960). At the extreme, it implies that only dissimilar individuals will coexist (``limiting similarity principle''; @MacArthur1967)\\

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.5\textwidth]{Figures/simulation_protocol.pdf}
\caption{\scriptsize{Illustration of the simulation protocol used in this paper.
A) we simulated a pure birth-tree with a trait evolving under a Brownian Motion process until reaching 200 tips resulting in a one dimensional trait space or two dimensional space.
We then applied different stressors to the resulting trait space using the following:
B) null mechanism: randomly removing half of the species resulting in two groups of 100 tips (blue and orange - the brown colour in the histogram represents the overlap between both groups);
C) equalising mechanism: removing species on the edge of the distribution;
D) facilitation mechanism: removing species to reduce the distance between pairs of species (increasing density);
E) filtering mechanism: removing species on one extreme of the distribution;
F) competition mechanism: proportionally removing species on the centre of the distribution (“flattening the curve”)..}}
\label{Fig:simulations}
\end{figure}
\bigskip

@@

We structured our simulations based on three aspects of diversity commonly captured by functional diversity metrics (@Mammola2021):
1) Regularity, encompassing metrics reflecting how regular the differences among observations are (4 metrics, equivalent to density metrics in @GuillermeEcoEvo2020);
2) Divergence, encompassing metrics reflecting the average difference among observations (6 metrics); and
3) Richness, encompassing metrics reflecting the sum of difference among observations; (4 metrics, equivalent to size metrics in @GuillermeEcoEvo2020?). Note that we focused here on three packages fully devoted to functional diversity analyses (BAT, FD, and TPD) in the R statistical environments (see @MammolaFuncEcol2021).

@@

\textbf{Aspect} & \textbf{Method} & \textbf{Name} & \textbf{Description} & \textbf{\texttt{R} function} & \textbf{Reference}\\
\hline
 & Trait Probability Density & Functional evenness & @@@NEEDS DESCRIPTION & \texttt{TPD::REND} & @Carmona2019\\
 & \cr 
 & Kernel density hypervolume & Functional evenness &  Overlap between a hypervolume and an imaginary hypervolume where traits are evenly distributed within their possible range & \texttt{BAT::kernel.evenness} & @MammolaCardoso2020\\
 & \cr 
 & Dendrogram & Functional evenness & Regularity of abundances and distances between observations in the tree. & \texttt{BAT::evenness} & @Cardoso2015\\
 & \cr 
 & Distances & Functional evenness & Regularity of abundances and distances along the minimum spanning tree linking all observations & \texttt{FD::dbFD()\$FEve} & @Villeger2008\\
\hline
Divergence & Trait Probability Density & Divergence & Overlap between the TPD function of a single observation and the TPD function of the whole set of observations & \texttt{TPD::REND} & Carmona2019\\
 & \cr 
 & Kernel density hypervolume & Functional dispersion  & Average distance between an observation and a sample of random points within and the probabilistic hypervolume & \texttt{BAT::kernel.dispersion} & @MammolaCardoso2020\\
 & \cr 
 & Dendrogram & Functional dispersion & Average dissimilarity between any two observations in the dendrogram & \texttt{BAT::dispersion} & @Cardoso2015\\
 & \cr  
 & Distances & Melodic Mean Pairwise Distance & @@@NEEDS DESCRIPTION  & melodic.rao & @Botta-Dukat2005\\
 & \cr 
 & Distances & Melodic Rao & @@@NEEDS DESCRIPTION & melodic.rao & @Botta-Dukat2005\\
 & \cr 
 & Distances & Rao's quadratic entropy & @@@NEEDS DESCRIPTION & \texttt{FD::dbFD()\$RaoQ} & Botta-Dukat2005\\
\hline
Richness & Trait Probability Density & Functional richness & Sum of cells where trait probability density $> 0$ & \texttt{TPD::REND} & @Carmona2019\\
 & \cr 
 & Kernel density hypervolume & Functional richness & Size (volume) of the hypervolume & \texttt{BAT::kernel.alpha} & @MammolaCardoso2020\\
 & \cr
 & Convex Hull & Functional richness & Volume of the convex hull & \texttt{BAT::hull.alpha} & @Cornwell2006\\
 & \cr
 & Dendrogram & Alpha diversity & Total branch length of the functional dendrogram & \texttt{BAT::alpha} & @PetcheyGaston2002; @Cardoso2015\\
\hline

@@

To ease interpretation of the results, we centred the results for the 14 metrics for the 4 non null stressors by the results for the null stressor and then scaled them.
Effectively, for each metric and each single simulation, we subtracted the results under the null stressor to the ones under the other stressors and divided the resulting centred metrics by the maximum value for each metric between each simulation.
This resulted in each metric being scaled between –1 and 1 where 0 corresponds to no change compared to the null, and 1 and –1 the maximum positive and negative deviations relative to the null.
In parallel, we also compared the distribution of the 200 scores for each metric with each reduction level for each non-null stressor to the null stressor (with no scaling or centering) using pairwise t-tests.
The results of these tests are displayed in table @@@ and in figure 2 (for the non-null stressor, grey distributions are not clearly distinguishable from the null stressor).

@@

We applied all metrics to an empirical dataset of bird extinctions on six Hawaiian Islands (Hawaii, Kauai, Lanai, Maui, Molokai, Oahu), an archipelago which has suffered large numbers of anthropogenic extinctions due to a range of extinction drivers @Walther2022
A prehistoric species list (avifauna known to be present prior to human colonisation of the islands), historic species list (avifauna known to be present at 1500 CE), and extant species list of the native community for each island was taken from @matthews along with trait values for the extinct species (Asio flammeus was removed from the historic and extant species lists as its colonisation status is uncertain)
We also added two marine species to the prehistoric species list; we compiled missing trait data for these two extinct species from @ferrans data
We extracted traits for extant species from the AVONET database (@tobias)
We used 9 morphological traits associated with dietary/foraging preference and dispersal ability @tobias @sheard to represent the functional diversity of the communities: mass; beak length (culmen); beak length (nares); beak width; beak depth; tarsus length; wing length; kipp’s distance (length from the first secondary feather to the tip of the longest primary feather @tobias); and tail length
Traits were log transformed and standardised to a mean of 0 and standard deviation of 1, prior to analyses.

The full trait dataset comprised 118 bird species, all native species known to have existed on these islands over the last 125,000 years
Fifty-five species are known to have gone extinct prior to 1500 CE, and 26 after 1500 CE
For our analyses here, we focused on two time periods: (i) the avifauna present in 1500 CE (the ``historic dataset'') and (ii) the current native avifauna (``extant dataset'')
We represented the distinct reductions in species richness preceding these two time periods as our two stressors, the first one representing all pre-1500 CE extinctions (i.e., $118-55$ species; a traitspace reduction of 46\%) and the second all the extinctions that have occurred until the present (i.e., $118-(55+26)$ species; a traitspace reduction of 69\%).

To build the community trait-space, we undertook a PCA including all 118 species, which we then subsetted to calculate the trait-space for the historic and extant datasets
We selected the first 5 axes to represent the trait-spaces that explained at least 97.5\% of the variance in each specific group (all, historic and extant species)
We applied the same procedure as for the simulated data by simulating a null mechanism to compare to the observed ones by randomly removing 55 and 81 species for each stressor and scaling the results proportionally to this null mechanism (as described above).

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.4\textwidth]{Figures/results_per_metric_per_stressor_emp.pdf}
\caption{\scriptsize{\textbf{Empirical results:} the metrics and the scaled changes are measured in the same way as in figure 2 (blue = regularity, yellow = divergence, green = richness).
Plain circles represent the metrics for the historic dataset (46\% removal), and fainted squares for the extant dataset (69\% removal).
}}
\label{Fig:simulation_results}
\end{figure}
\bigskip

>>>>

\documentclass[12pt,letterpaper]{article}
\usepackage{natbib}

@@

In the last two decades, there has been a progressive shift in ecology and evolution from  taxonomic-oriented approaches, with species as the focal point of studies, to functional approaches that place species-specific characteristics (traits) at the centre of analyses \citep{mammola2021concepts, palacio2022protocol}).
Although numerous definitions exist \citep{dawson2021traits}, we here consider a trait to be any characteristic - e.g. , morphological, anatomical, ecological, physiological, behavioural, phenological—measured on individual organisms at any level, from genes to whole organisms.
A trait-based approach has two advantages for answering core questions in ecology and evolution.
First, it allows a deeper understanding of the mechanisms generating biodiversity patterns, by putting organisms’ traits at the centre of natural selection rather than organisms themselves.
Second, it allows comparisons across subdisciplines in biology, while facilitating the conceptualization of general principles broadly valid in space (e.g. unrelated species pools) and time (e.g. anatomical traits are comparable between palaeontology and ecology, which is not always the case with species; \citealt{luza2023going}).
Already used routinely in palaeontology \cite{raup1961geometry, gould1991disparity, foote1995morphological, guillerme2020disparities}, this new trait-focused ecology and evolution is unlocking the possibility to answer a broad range of questions in disciplines as diverse as community ecology \cite{mcgill2006rebuilding}, biogeography \cite{violle2014emergence}, conservation biology \cite{chichorro2022trait}, micro- \cite{chapin1993evolution}, macro-evolution \cite{wiens2023trait} %TG: check to find another cite
, and applied fields (e.g. agronomy \citealt{martin2015plant}).
This is because trait-based approaches closely align with the general analytical framework proposed by \cite{anand1994pattern} to answer three consequential questions: \textit{``what?''} (describing the pattern), \textit{``how?''} (describing the process), and \textit{``why?''} (understanding the mechanism) (Box 1).

\bigskip
\bigskip
\hrule
\hrule

\section*{Box}
\cite{anand1994pattern} conceptualised a general analytical framework for answering scientific questions in ecology and evolution, based on three consequential questions: \textit{``what?''} (describing the pattern), \textit{``how?''} (describing the process), and \textit{``why?''} (understanding the mechanism)
This framework can be effectively applied to trait-based analyses:

\textit{What?} Pattern description corresponds to the steps needed to collect and summarise data to answer a question of interest (i.e., ``how?'' and ``why?'' as defined below).
Usually this consists in: i) collecting target traits at the focal level (e.g., genes, individual, population, species; \citealt{violle2007let}); ii) arranging these traits into some kind of trait space \citep{guillerme2020disparities, mammola2021concepts}; and iii) using one or more unidimensional statistics to summarise properties of the trait space.
For unidimensional spaces (i.e. distributions), these statistics can be as simple as the mean and the standard deviation (e.g. community weighted mean).
For multidimensional spaces, these are usually named disparity metrics or indices \citep{guillerme2020disparities} or functional diversity metrics \citep{mammola2021concepts}, but all attempt to capture some pattern of interest in the trait space.

\textit{How?} Process description can be seen as linking the pattern of interest (``What?'') to some dynamic element.
This can be a punctual change.
For example, change of the pattern under a certain condition, such as how traits X differ between two habitats \citep{martinez2021habitat}.
But it can also be one or more continuous or ordinal changes, such as how the pattern X changes in space, time and/or along ecological gradients \citep{belmaker2013spatial, jarzyna2018taxonomic,lamanna2014functional,bjorkman2018plant,mclean2021trait}.
The distinction here might seem trivial but it is significant: usually, the process designates the change of the pattern, not the change of the traits.
Although the traits are what is really changing, researchers will usually analyse some emergent property of the trait aggregation as described by the change of statistics between two or more conditions.

\textit{Why?} Mechanism description is then at the core of answering the biological question at hand.
In most cases, this is what researchers are actually trying to understand.
For example, one might be interested in understanding the effect of climate change on some trait (e.g. Does temperature affect leaf surface area? @@@). %TG: to check
In fact, most researchers work on understanding the causal link between variables (i.e. ``why'' are ``what'' and ``how'' linked).
This can be very useful, for example, to provide predictions about the past or the future.
Ultimately, studying the mechanisms (``why'') is often the reason why researchers get funded (or not).

@@

\bigskip
\hrule
\hrule
\bigskip

While the first part of this framework (``what?'') has been studied extensively in the last decade (@CITECITECITE), %TG: TODO
 this was often undertaken outside of the ``what, how, and why'' framework.
Here we argue that the pattern description, although the basis of any future study, often follows some framework without a proper evaluation of its adequacy to answer the how and why.
The choice of the tool or metric used to describe the pattern is crucial for allowing us to understand the process and the mechanism.
Using an inappropriate metric for describing a pattern can lead to biased conclusions.
For example, let us imagine one is interested in understanding how two populations compete with each other for a resource (the mechanism - ``why?'') on an island compared to another island with only one of the species (the process - ``how?'').
In such a scenario, measuring the occupied area of each population on each island (a pattern - ``what?'') will not be the most appropriate way to understand the potential competition between these populations.
In this simplistic example, the functional overlap between populations (``what and how'') might be more appropriate.
See for example \citealt{carvalho2020decomposing} for a discussion of how Darwin’s finches share resources depending on the existence or not of competition.

Through a simulation exercise, we analysed different patterns (what) across different processes (how) approximating different mechanisms (why) of interaction between organisms: equalisation, filtering, facilitation and competition (Figure \ref{Fig:simulations}).
We also used an empirical dataset of Hawaiian bird traits and used pre-historical and historical extinction events on Hawaiian islands as patterns and processes capturing the effects of species extinction on species trait distributions.
We show that, with a fixed process and pattern, the choice of statistics used to describe the patterns (the disparity or functional diversity metrics) have a great impact on our ability to recover the process and understand the mechanism.
We suggest caution when summarising patterns in observed data and propose tools for researchers to help understand how their pattern (often stemming from multidimensional data) can vary intuitively or not depending on the process and mechanism of interest.

@@

First, we simulated a single Brownian Motion trait under a pure birth speciation model until reaching 200 elements (representing tips, individuals, species, OTUs, etc.) in \texttt{R} \citep{rcore} using \texttt{treats} \citep{guillerme2024treats}.
This resulted in a neutral null model of trait evolution with no effect of competition, extinction, selection or other processes (\textit{sensu} \citealt{bausman2018modeling}).
We refer to this as the ``non-stressed trait space''.

@@

We then applied five different stressors to the non-stressed trait space to remove either 20\%, 40\%, 60\% or 80\% of the data (resulting in trait spaces with, respectively, 160, 120, 80 and 40 elements).
In removing the data, we applied these five specific algorithms (Figure \ref{Fig:simulations}; Table \ref{Tab:mechanisms}; all algorithms, except ``Evenness'', were previously described in \cite{guillerme2020shifting}):

@@

For the position, size and density algorithms, we estimated the parameters $\rho$ and $D$ recursively with the \texttt{dispRity} package to obtain the required amount of data to be removed (\texttt{dispRity::reduce.space}, \citealt{guillerme2018disprity,guillerme2020shifting}).
These steps resulted in 4000 simulated trait spaces (200 simulated spaces $\times$ 5 stressors $\times$ 4 amounts of data removed).
See Table \ref{Tab:mechanisms} for a biological description of these stressors and Figure \ref{Fig:simulations} for a visual description of them.

@@

We repeated each simulation pipeline (generate a trait space and apply the stressor) for uncorrelated 4 dimensional traits (Figure \ref{Fig:simulation_results}) and uncorrelated 2, 3, 5 and 8 dimensional traits (supplementary materials).
We limited our simulations to a relatively small number of traits due to the constraints of some of the metrics used (e.g., \texttt{TPD::TPDsMean} is only implemented for up to 4 dimensions; \citealt{carmona2019trait}) but also to avoid dealing with the curse of dimensionality \citep{bellman1957dynamic}.
This curse changes the properties of space occupancy in a non-linear way depending on the element's distribution and the number of dimensions.
For example, the volume of a trait space (or hypervolume when using more than 3 dimensions) typically tends to zero in a high number of dimensions.
However, the rate at which it approaches zero is not linear and depends on the distribution of each element on every dimension.
This makes it practically difficult to compare two randomly generated spaces with similar characteristics (e.g. for two spaces with 200 elements and 10 dimensions generated in the exact same way, one might have a hypervolume of nearly 0 and the other one of $10^5$).
Note also that in our simulations, all dimensions have the same properties (i.e.
same variance and distribution).
This is often not the case in empirical cases (e.g. see our empirical example) where the dimensions have a decreased variance due to ordination techniques.

\begin{table}
\center
\scriptsize
\begin{tabular}{p{0.3\linewidth}|p{0.3\linewidth}|p{0.4\linewidth}}
\textbf{Approximated mechanism (biological)} & \textbf{Used mechanism (algorithm in \texttt{dispRity::reduce.space})} & \textbf{Stressor description}\\
\hline
  Null mechanism & Random removal (``random'') & No overall change in species traits or community structure except for the reduction in the number of elements.\\
  Equalising fitness & Size change (``size'') & When species with trait combinations that are more extreme (i.e. species located away from the centre of the trait space) are disadvantaged due to some resource concentration gradient. \\
  Facilitation & Density change (``density'') & When species with closely shared trait combinations (i.e. located close together in the trait space) are more likely to survive the stressor. In other words, this simulates the idea that some species are more likely to be resilient if they are present in a community of species with shared traits rather than the opposite.\\
  Environmental filtering & Position change (``position'') & When an increase in species trait similarity happens through some strong abiotic constraints \citep{cornwell2006trait}. For example, some trait combinations become more and more unlikely due to some environmental constraints (i.e. some regions of the trait space become unsuitable).\\
  Competitive exclusion & Evenness change (``evenness'') & When functionally similar species compete more intensively with one another than with functionally dissimilar species (``competitive exclusion principle''; \citealt{hardin1960competitive}). At the extreme, it implies that only dissimilar individuals will coexist (``limiting similarity principle''; \citealt{macarthur1967limiting})\\

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.5\textwidth]{Figures/simulation_protocol.pdf}
\caption{\scriptsize{Illustration of the simulation protocol used in this paper.
A) we simulated a pure birth-tree with a trait evolving under a Brownian Motion process until reaching 200 tips resulting in a one dimensional trait space or two dimensional space.
We then applied different stressors to the resulting trait space using the following:
B) null mechanism: randomly removing half of the species resulting in two groups of 100 tips (blue and orange - the brown colour in the histogram represents the overlap between both groups);
C) equalising mechanism: removing species on the edge of the distribution;
D) facilitation mechanism: removing species to reduce the distance between pairs of species (increasing density);
E) filtering mechanism: removing species on one extreme of the distribution;
F) competition mechanism: proportionally removing species on the centre of the distribution (``flattening the curve'')..}}
\label{Fig:simulations}
\end{figure}
\bigskip

@@

We structured our simulations based on three aspects of diversity commonly captured by functional diversity metrics \citep{mammola2021concepts}:
1) Regularity, encompassing metrics reflecting how regular the differences among observations are (4 metrics, equivalent to density metrics in \citealt{guillerme2020shifting});
2) Divergence, encompassing metrics reflecting the average difference among observations (6 metrics); and
3) Richness, encompassing metrics reflecting the sum of difference among observations; (4 metrics, equivalent to size metrics in @GuillermeEcoEvo2020?). Note that we focused here on three packages fully devoted to functional diversity analyses (BAT, FD, and TPD) in the R statistical environments (see \citealt{mammola2021concepts}).

@@

\textbf{Aspect} & \textbf{Method} & \textbf{Name} & \textbf{Description} & \textbf{\texttt{R} function} & \textbf{Reference}\\
\hline
 & Trait Probability Density & Functional evenness & @@@NEEDS DESCRIPTION & \texttt{TPD::REND} & \cite{carmona2019trait}\\
 & \cr 
 & Kernel density hypervolume & Functional evenness &  Overlap between a hypervolume and an imaginary hypervolume where traits are evenly distributed within their possible range & \texttt{BAT::kernel.evenness} & \cite{mammola2020functional}\\
 & \cr 
 & Dendrogram & Functional evenness & Regularity of abundances and distances between observations in the tree. & \texttt{BAT::evenness} & \cite{cardoso2015bat}\\
 & \cr 
 & Distances & Functional evenness & Regularity of abundances and distances along the minimum spanning tree linking all observations & \texttt{FD::dbFD()\$FEve} & \cite{Laliberte2010FD,villeger2008new}\\
\hline
Divergence & Trait Probability Density & Divergence & Overlap between the TPD function of a single observation and the TPD function of the whole set of observations & \texttt{TPD::REND} & \cite{carmona2019trait}\\
 & \cr 
 & Kernel density hypervolume & Functional dispersion  & Average distance between an observation and a sample of random points within and the probabilistic hypervolume & \texttt{BAT::kernel.dispersion} & \cite{mammola2020functional}\\
 & \cr 
 & Dendrogram & Functional dispersion & Average dissimilarity between any two observations in the dendrogram & \texttt{BAT::dispersion} & \cite{cardoso2015bat}\\
 & \cr  
 & Distances & Melodic Mean Pairwise Distance & @@@NEEDS DESCRIPTION  & melodic.rao & \cite{botta2005rao}\\
 & \cr 
 & Distances & Melodic Rao & @@@NEEDS DESCRIPTION & melodic.rao & @Botta-Dukat2005\\
 & \cr 
 & Distances & Rao's quadratic entropy & @@@NEEDS DESCRIPTION & \texttt{FD::dbFD()\$RaoQ} & \cite{botta2005rao}\\
\hline
Richness & Trait Probability Density & Functional richness & Sum of cells where trait probability density $> 0$ & \texttt{TPD::REND} & \cite{carmona2019trait}\\
 & \cr 
 & Kernel density hypervolume & Functional richness & Size (volume) of the hypervolume & \texttt{BAT::kernel.alpha} & \cite{mammola2020functional}\\
 & \cr
 & Convex Hull & Functional richness & Volume of the convex hull & \texttt{BAT::hull.alpha} & \cite{cornwell2006trait}\\
 & \cr
 & Dendrogram & Alpha diversity & Total branch length of the functional dendrogram & \texttt{BAT::alpha} & \cite{petchey2002functional,cardoso2015bat}\\
\hline

@@

To ease interpretation of the results, we centred the results for the 14 metrics for the 4 non null stressors by the results for the null stressor and then scaled them.
Effectively, for each metric and each single simulation, we subtracted the results under the null stressor to the ones under the other stressors and divided the resulting centred metrics by the maximum value for each metric between each simulation.
This resulted in each metric being scaled between –1 and 1 where 0 corresponds to no change compared to the null, and 1 and –1 the maximum positive and negative deviations relative to the null.
In parallel, we also compared the distribution of the 200 scores for each metric with each reduction level for each non-null stressor to the null stressor (with no scaling or centering) using pairwise t-tests.
The results of these tests are displayed in table @@@ %TG: TODO: add table
 and in figure \ref{Fig:simulation_results} (for the non-null stressor, grey distributions are not clearly distinguishable from the null stressor).

@@

We applied all metrics to an empirical dataset of bird extinctions on six Hawaiian Islands (Hawaii, Kauai, Lanai, Maui, Molokai, Oahu), an archipelago which has suffered large numbers of anthropogenic extinctions due to a range of extinction drivers \citep{Walther2022}.
A prehistoric species list (avifauna known to be present prior to human colonisation of the islands), historic species list (avifauna known to be present at 1500 CE), and extant species list of the native community for each island was taken from \cite{matthews2023global} along with trait values for the extinct species (Asio flammeus was removed from the historic and extant species lists as its colonisation status is uncertain).
We also added two marine species to the prehistoric species list; we compiled missing trait data for these two extinct species from @ferrans data
We extracted traits for extant species from the AVONET database \citep{tobias2022avonet}
We used 9 morphological traits associated with dietary/foraging preference and dispersal ability \citep{pigot2020macroevolutionary,sheard2020ecological} to represent the functional diversity of the communities: mass; beak length (culmen); beak length (nares); beak width; beak depth; tarsus length; wing length; kipp’s distance (length from the first secondary feather to the tip of the longest primary feather; \citealt{tobias2022avonet}); and tail length
Traits were log transformed and standardised to a mean of 0 and standard deviation of 1, prior to analyses.

The full trait dataset comprised 118 bird species, all native species known to have existed on these islands over the last 125,000 years.
Fifty-five species are known to have gone extinct prior to 1500 CE, and 26 after 1500 CE.
For our analyses here, we focused on two time periods: (i) the avifauna present in 1500 CE (the ``historic dataset'') and (ii) the current native avifauna (``extant dataset'').
We represented the distinct reductions in species richness preceding these two time periods as our two stressors, the first one representing all pre-1500 CE extinctions (i.e., $118-55$ species; a trait space reduction of 46\%) and the second all the extinctions that have occurred until the present (i.e., $118-(55+26)$ species; a trait space reduction of 69\%).

To build the community trait-space, we undertook a PCA including all 118 species, which we then subsetted to calculate the trait-space for the historic and extant datasets.
We selected the first 5 axes to represent the trait-spaces that explained at least 97.5\% of the variance in each specific group (all, historic and extant species)
We applied the same procedure as for the simulated data by simulating a null mechanism to compare to the observed ones by randomly removing 55 and 81 species for each stressor and scaling the results proportionally to this null mechanism (as described above).

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.4\textwidth]{Figures/results_per_metric_per_stressor_emp.pdf}
\caption{\scriptsize{\textbf{Empirical results:} the metrics and the scaled changes are measured in the same way as in figure 2 (blue = regularity, yellow = divergence, green = richness).
Plain circles represent the metrics for the historic dataset (46\% removal), and fainted squares for the extant dataset (69\% removal).
}}
\label{Fig:empirical_results}
\end{figure}
\bigskip
